Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-17 paper/ms.tex
wordsmithing

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is probably caused by eclipses of dust or gas clumps orbiting at
  the Keplerian corotation radius.  Here, we report new CQVs
  discovered in TESS short-cadence data collected between 2018~July
  and 2022~September.  Our search of \nstarssearched\ K and M dwarfs with
  $T$$<$16 and $d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most
  of these discoveries are new, and they include the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) examples of this class of object currently
  known.  A few objects are outliers: LP 12-502 for instance showed a
  ``dip complex'' with a period and duration that were fixed over
  1,500 cycles, but in detail, this system exhibited between four and
  eight local minima per cycle, and at times also displayed multiple
  periods simultaneously.  We demonstrate that transient corotating
  material is the most viable explanation for this class of object,
  and present evidence supporting the hypothesis that this material is
  sculpted by stellar magnetic fields dominated by high order
  multipoles.  We expect that our sample will facilitate future
  efforts aimed at connecting CQVs to the broader contexts of star,
  disk, and perhaps even exoplanet evolution. 	
\end{abstract}

@@

A point on nomenclature.  CQVs have been called ``transient flux
dips'', ``persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and which show large-amplitude
variability linked to obscuring inner disk structures and accretion
hot spots \citep{2014AJ....147...82C,2021ApJ...908...16R}.  While a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}), their
phenomenology and stellar properties are quite different.  At the risk
of introducing yet another standard, we hope to use a nomenclature
that reflects how, when observed over timescales of more than tens of
cycles, CQVs are almost but not exactly periodic.  They are
quasiperiodic.  While the three-type classification scheme proposed by
\citet{2017AJ....153..152S} may indeed provide some helpful visual
distinctions amongst the CQVs, it seems likely that they are all
explained by a single underlying phenomenon, and so we opt to refer to
them by a single empirically descriptive name.  We also considered
names that focused on the presence of transient corotating clumps, or
more generally on circumstellar plasma tori; for the time being,
``CQV'' is agnostic to the true underlying phenomenon.

@@

To construct our sample, we analyzed the short-cadence data acquired
by TESS between 2018 July 25 and 2022 September 1 (Sectors 1-55).
Specifically, we used the 120-second cadence light curves produced by
the Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products from these
sectors also included full frame images with cadences of 600 and 1800
seconds, we limited our scope in this study for the sake of simplicity
in data handling and some degree of uniformity in the properties of
the data itself.  In exchange, we sacrificed both completeness and
homogeneity of the selection function.  While TESS cumulatively
observed $\approx$90\% of the sky for at least one lunar month between
2018 July and 2022 September, the 120-second cadence data were
collected for only a subset of observable stars during this time due
to telemetric constraints \citep[see][]{2021PASP..133i5002F}.

To assess the completeness of the resulting 120-second cadence data
that is the basis of this study, we cross-matched TIC8.2
\citep{2019AJ....158..138S,2021arXiv210804778P} against the Gaia DR2 point-source catalog
\citep{2018A&A...616A...1G}.  We opted for Gaia DR2 rather than DR3
because the base catalog for TIC8 was Gaia DR2, which facilitated a
one-to-one crossmatch using the Gaia source identifiers.  This
exercise showed that for $T$$<$16 M dwarfs, the TESS 2-minute data are
$\approx$50\% complete at $\approx$50\,pc.  At $<$20\,pc,
$\gtrsim$80\% of the $T$$<$16 M dwarfs have at least one sector of
short-cadence data; at $>$100\,pc, $\lesssim$10\% of such M dwarfs
have at least one sector of short-cadence data.  Armed with this
understanding, we then used our cross-match between Gaia DR2 and TIC8
to select our stars of interest, which we defined as stars with
120-second cadence TESS light curves satisfying these four conditions:
\begin{align}
  T &< 16 \quad&(\mathrm{Bright\ for\ TESS})\\
  \bprp &> 1.5 \quad&(\mathrm{Red\ stars\ only})\\
  M_{\rm G} &> 4 \quad&(\mathrm{Dwarf\ stars\ only})\\
  d &< 150\,{\rm pc} \quad&(\mathrm{Close\ stars\ only}).
\end{align}
Here, $M_{\rm G} = G + 5\log(\varpi_{\rm as}) + 5$ is the Gaia
$G$-band absolute magnitude, $\varpi_{\rm as}$ is the parallax in
units of arcseconds, and $d$ is a geometric distance defined by
inverting the parallax and ignoring any zero-point correction.  The
target sample therefore includes \nstarssearched\ M dwarfs and late-K
dwarfs, down to $T$$<$16 and out to $d$$<$150\,pc.

@@

Previous methods for finding CQVs have included visually examining
stars known to be in young clusters
\citep{2016AJ....152..114R,2017AJ....153..152S}, and automatically
flagging rapid rotators with a large number of strong Fourier
harmonics \citep{2019ApJ...876..127Z}.  The latter approach still
requires visual vetting, since ``stars with many Fourier harmonics''
is a designation that includes objects such as eclipsing binaries or
multiple stars blended into a single photometric aperture.  In this
work, we implemented a new search approach based on counting the
number of sharp local minima in phase-folded light curves, while also
using the previously tested Fourier approach.  We applied these two
search techniques independently.

@@

We acquired iodine-free reconnaissance spectra using Keck/HIRES for
three CQVs.  The goals were to determine the chromospheric activity
levels, and to check for indications of either accretion or
spectroscopic binarity.  We acquired a 15 minute exposure of
TIC~146539195 on 2023 January 3, a 15 minute exposure of TIC~264599508
on 2023 January 9, and a 30 minute exposure of TIC~402980664 on 2023
July 10.  The acquisition and analysis followed the usual techniques
of the California Planet Survey \citep{2010ApJ...721.1467H}.
Figure~\ref{fig:speccutouts} shows cutouts from the resulting spectra,
centered on the \ion{Ca}{2} HK windows, H$\alpha$, and the \ion{Li}{1}
6708\,\AA\ doublet.  The \ion{Ca}{2} H emission line is blended with
H$\epsilon$.  While a more detailed analysis will be left for future
work, these spectra confirm previous understanding established by
\citet{2017AJ....153..152S} that the stars are chromospherically
active M dwarfs in the ``weak-lined'' T Tauri regime
\citep[e.g.][Figure~15]{2019AJ....157...85B}.  Their H$\alpha$
equivalent widths, at $\approx$14\,\AA, $\approx$8\,\AA, and
$\approx$3\AA, are consistent with purely chromospheric emission.
The blue excess in TIC~264599508 could be explained by a second
unresolved star; the TESS light curve for this source shows both the
7.90\,hr CQV signal, and a 9.00\,hr rotation signal with comparable
amplitude.

>>>>

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is probably caused by eclipsing clumps of dust or gas
  in orbit at the Keplerian corotation radius.  Here, we
  report new CQVs discovered in TESS short-cadence data collected
  between 2018~July and 2022~September.  Our search of
  \nstarssearched\ K and M dwarfs with $T$$<$16 and $d$$<$150\,pc
  yielded \ngoods\ high-quality CQVs.  Most of these discoveries are
  new, and they include the brightest ($T$$\approx$9.5), closest
  ($d$$\approx$20\,pc), and oldest ($\approx$200\,Myr) examples of
  this class of object currently known.  A few objects are outliers:
  LP 12-502 for instance showed a ``dip complex'' with a period and
  duration that were fixed over 1,500 cycles, but in detail, this
  system exhibited between four and eight local minima per cycle, and
  at times also displayed multiple periods simultaneously.  We
  demonstrate that transient corotating material is the most viable
  explanation for this class of object, and present evidence
  supporting the hypothesis that this material is sculpted by stellar
  magnetic fields dominated by high order multipoles.  We expect that
  our sample will facilitate future efforts aimed at connecting CQVs
  to the broader contexts of star, disk, and perhaps even exoplanet
  evolution. 	
\end{abstract}

@@

A point on nomenclature.  CQVs have been called ``transient flux
dips'', ``persistent flux dips'', ``scallop shells'', ``batwings'',
\citep{2017AJ....153..152S} ``complex rotators'',
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
and ``complex periodic variables'' \citep{2023MNRAS.518.2921K}.  The
CQVs should not be conflated with ``dippers'', which are classical T
Tauri stars with infrared excesses, and which show large-amplitude
variability linked to obscuring inner disk structures and accretion
hot spots \citep{2014AJ....147...82C,2021ApJ...908...16R}.  While a
few similarities between CQVs and dippers do exist (see
Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}), their
phenomenology and stellar properties are quite different.  At the risk
of introducing yet another standard, we hope to use a nomenclature
that reflects how, when observed over timescales of more than tens of
cycles, CQVs are almost but not exactly periodic.  They are
quasiperiodic.  While the three-type classification scheme proposed by
\citet{2017AJ....153..152S} may provide helpful visual
distinctions amongst the CQVs, it seems likely that they are all
explained by a single underlying phenomenon, and so we opt to refer to
them by a single empirically descriptive name.  We also considered
names that focused on the presence of transient corotating clumps, or
more generally on circumstellar plasma tori; for the time being,
``CQV'' is agnostic to the true underlying phenomenon.

@@

To construct our sample, we analyzed the short-cadence data acquired
by TESS between 2018 July 25 and 2022 September 1 (Sectors 1-55).
Specifically, we used the 120-second cadence light curves produced by
the Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products from these
sectors also included full frame images with cadences of 600 and 1800
seconds, we limited our scope in this study for the sake of simplicity
in data handling and uniformity in the data.  In exchange, we
sacrificed both completeness and homogeneity of the selection
function.  While TESS cumulatively observed $\approx$90\% of the sky
for at least one lunar month between 2018 July and 2022 September, the
120-second cadence data were collected for only a subset of observable
stars during this time due to telemetric constraints
\citep[see][]{2021PASP..133i5002F}.

To assess the completeness of the resulting 120-second cadence data
that are the basis of this study, we cross-matched TIC8.2
\citep{2019AJ....158..138S,2021arXiv210804778P} against the Gaia DR2 point-source catalog
\citep{2018A&A...616A...1G}.  We opted for Gaia DR2 rather than DR3
because the base catalog for TIC8 was Gaia DR2, which facilitated a
one-to-one crossmatch using the Gaia source identifiers.  This
exercise showed that for $T$$<$16 M dwarfs, the TESS 2-minute data are
$\approx$50\% complete at $\approx$50\,pc.  At $<$20\,pc,
$\gtrsim$80\% of the $T$$<$16 M dwarfs have at least one sector of
short-cadence data; at $>$100\,pc, $\lesssim$10\% of such M dwarfs
have at least one sector of short-cadence data.  Armed with this
understanding, we then used our cross-match between Gaia DR2 and TIC8
to select our stars of interest, which we defined as stars with
120-second cadence TESS light curves satisfying these four conditions:
\begin{align}
  T &< 16 \quad&(\mathrm{Bright\ for\ TESS})\\
  \bprp &> 1.5 \quad&(\mathrm{Red\ stars\ only})\\
  M_{\rm G} &> 4 \quad&(\mathrm{Dwarf\ stars\ only})\\
  d &< 150\,{\rm pc} \quad&(\mathrm{Close\ stars\ only}).
\end{align}
Here, $M_{\rm G} = G + 5\log(\varpi_{\rm as}) + 5$ is the Gaia
$G$-band absolute magnitude, $\varpi_{\rm as}$ is the parallax in
units of arcseconds, and $d$ is a geometric distance defined by
inverting the parallax and ignoring any zero-point correction.  The
target sample therefore includes \nstarssearched\ M dwarfs and late-K
dwarfs, down to $T$$<$16 and out to $d$$<$150\,pc.

@@

Previous methods for finding CQVs have included visually examining
stars in young clusters
\citep{2016AJ....152..114R,2017AJ....153..152S} and automatically
flagging rapid rotators with a large number of strong Fourier
harmonics \citep{2019ApJ...876..127Z}.  The latter approach still
requires visual vetting, since ``stars with strong Fourier harmonics''
is a designation that includes objects such as eclipsing binaries or
multiple stars blended into a single photometric aperture.  In this
work, we implemented a new search approach based on counting the
number of sharp local minima in phase-folded light curves, while also
using the Fourier approach.  We applied these two
search techniques independently.

@@

We acquired iodine-free reconnaissance spectra using Keck/HIRES for
three CQVs.  The goals were to determine the chromospheric activity
levels, and to check for indications of either accretion or
spectroscopic binarity.  We acquired a 15 minute exposure of
TIC~146539195 on 2023 January 3, a 15 minute exposure of TIC~264599508
on 2023 January 9, and a 30 minute exposure of TIC~402980664 on 2023
July 10.  The acquisition and analysis followed the usual techniques
of the California Planet Survey \citep{2010ApJ...721.1467H}.
Figure~\ref{fig:speccutouts} shows cutouts from the resulting spectra,
centered on the \ion{Ca}{2} HK windows, H$\alpha$, and the \ion{Li}{1}
6708\,\AA\ doublet.  The \ion{Ca}{2} H emission line is blended with
H$\epsilon$.  While a more detailed analysis will be left for future
work, these spectra confirm previous understanding established by
\citet{2017AJ....153..152S} that the stars are chromospherically
active M dwarfs in the ``weak-lined'' T Tauri regime
\citep[e.g.][Figure~15]{2019AJ....157...85B}.  Their H$\alpha$
equivalent widths, at $\approx$14\,\AA, $\approx$3\AA, and
$\approx$8\,\AA (for TIC~264599508, 146539195, and 402980664
respectively) are consistent with purely chromospheric emission.  The
blue excess in TIC~264599508 could be explained by a second unresolved
star; the TESS light curve for this source shows both the 7.90\,hr CQV
signal, and a 9.00\,hr rotation signal with comparable amplitude.
